---
title: "Investigating COVID-19 Virus Trends"
author: "Abe Ceasar Perez"
date: "March 13, 2022"
output: html_document
---
<br>

### Understanding the Impact of COVID-19 around the Globe

This project aims to identify which countries have had the highest number of positive cases against the number of tests. We will be working for an unnamed health organization to determine which areas around the globe have been badly hit by the COVID-19 virus, an airborne virus which originated from Wuhan, China.
<br>
<br>
For this project, we will be using basic concepts regarding data structures such as vectors, lists, dataframes, and matrices in order to achieve the objective above.
<br>

### Initial Findings

Based on the outcome of this project, we have identified three countries that have the highest positive-to-test ratio and these are: United Kingdom, United States, and Turkey.
<br>
<br>
The process for these results are summarized below:

### Inspecting the COVID-19 Dataset

We will be using this dataset from Kaggle, a popular website for practitioners and data enthusiasts who wanted to use open-sourced datasets: <https://www.kaggle.com/lin0li/covid19testing>.
<br>
<br>
We will be loading the necessary packages for our analysis along with the dataset to be used for this project.

```{r load_dataset, include=TRUE, warning=FALSE}

# loads packages without warnings/messages
suppressPackageStartupMessages(library(readr)) # for reading the data
suppressPackageStartupMessages(library(tibble)) # for creating tibbles
suppressPackageStartupMessages(library(dplyr)) # for loading/manipulating functions

covid_df <- read_csv("covid19.csv", show_col_types = FALSE) # reads data without showing the column names
head(covid_df) # displays the first few rows
```

For documentation purposes later, we will be storing the column names into a character vector. The structure of the object is also displayed as demonstrated below.

```{r inspect_cols, include=TRUE}

vector_cols <- colnames(covid_df)
vector_cols # object created is a character vector

```

Looking further into the dataset, we can see that the dataframe has nearly 11,000 rows and 14 columns. The data that we'll be using was collected on a daily basis and were labeled per continent all the way to the province/state level. Upon further inspection, we can see that there is also aggregated data on some regions which are mixed in with the data for each individual region.

```{r inspect_df, include=TRUE}

glimpse(covid_df) # gives a complete overview of what the entire dataset contains

```

### Analyzing country-level data

Since our analysis focuses on data collected per country, we will be filtering out rows in which the province/state column are set to "All states" to only show the aggregated data per country. Afterwards, we will create another dataframe which includes only the columns relevant to the analysis. 

Since the dataframe also contains cumulative data, we will be excluding those variables from the analysis since our focus will be on the daily numbers.

```{r filter_df, include=TRUE}

covid_df_all_states <- covid_df %>% 
  filter(Province_State ==  "All States") %>% 
  select(-Province_State) # we can remove province_state since we have already filtered out our data to all states

covid_df_all_states_daily <- covid_df_all_states %>% 
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive) # Select columns related to daily measures

```

After the dataframe has been filtered out, we will be combining the daily measures by country and arrange the measures by total cases in order to determine the top 10 countries with the most number of cases.

```{r summarize, include=TRUE}

covid_df_all_states_daily_sum <- covid_df_all_states_daily %>% 
  group_by(Country_Region) %>% 
  summarize(tested = sum(daily_tested), 
            positive = sum(daily_positive), 
            active = sum(active),
            hospitalized = sum(hospitalizedCurr)) %>% 
  arrange(desc(tested)) # Aggregate daily data per country

covid_df_all_states_daily_sum

covid_top_10 <- head(covid_df_all_states_daily_sum, 10) ##Stores first 10 rows

```

We will then store the relevant data for each column into vectors and label them by country for ease of analysis.

```{r vectorize, include=TRUE}

# Create vectors for country names and for each daily measure from the top 10 dataframe
# Each measure is then labeled by their corresponding country name

countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized

names(tested_cases) <- countries
names(positive_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries

```

Using the created vectors above, we will now calculate the positive to tested cases ratio by dividing the total number of positives over the total number of tested cases. We will also arrange the vector to determine the top 3 countries with the highest ratios.

```{r identify_top_3, include=TRUE}

# Calculate the positives-to-tests ratio per country
positive_cases/tested_cases
# Outputs the top 3 countries by ratio
positive_tested_top_3 <- round(sort(positive_cases/tested_cases, 
                                    decreasing = TRUE),4)[1:3]
positive_tested_top_3 

```

### Consolidating all the Findings

After identifying the top 3 countries, we will be creating a matrix to store the aggregated data for each country including the ratios computed above.

```{r convert_to_matrix, include=TRUE}

# creates vectors for the top countries using the values from the top 10 dataframe including the calculated ratio from the previous code
united_kingdom <- c(0.11, 1473672, 166909, 0, 0) 
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

covid_mat <- rbind(united_kingdom, united_states, turkey)
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")
covid_mat

```

Since we have now identified the countries of interest, we will be storing all of our findings on a structured list which contains the list of all the data structures used as well as the question and answer to our research problem.

```{r create_lists, include=TRUE}

##Stores all findings in a structured list

question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)

dataframes <- list(data = covid_df,
                   data_by_country = covid_df_all_states,
                   daily_country_data = covid_df_all_states_daily,
                   top_10_countries_by_cases = covid_top_10) # creates a list of all dataframes used
matrices <- list(top_3_country_stats = covid_mat) # creates a list of all matrices used
vectors <- list(data_cols = vector_cols, country_names = countries) # creates a list of all vectors used

data_structure_list <- list(dataframes = dataframes, 
                            matrices = matrices,
                            vectors = vectors) # combines all lists into one structured list

covid_analysis_list <- list(question, answer, data_structure_list) # consolidates analysis into one master list

```

### Which countries are highly impacted by COVID-19?

Based on the methods above, we have identified three countries with the highest positivity/case ratios namely, United Kingdom, United States, and Turkey. 

Although United Kingdom had the highest positivity ratio, the United States remains to have the highest number of cases accumulated throughout the period, with over 10 times compared to the two countries mentioned.
 
```{r answer, include=TRUE}

## Here are the top 3 countries with the highest positive to tested cases ratio.
covid_analysis_list[[2]] 

```

### Conclusion

In this project, we were able to determine the top three countries with the highest impact from the COVID-19 Pandemic. We were able to apply the concepts of data structures in order to analyze the data and organize our results in a more cohesive manner.
<br>
<br>
We were able to determine that United Kingdom experienced the highest impact on the COVID-19 pandemic despite the United States having the highest amount of cases. We will then recommend the organization to urgently monitor the countries mentioned above and advise these countries to further implement stricter measures in order to control the spread of the virus.
<br>
<br>